---
title: "Lubridate Package"
author: "Miraj Patel"
date: "2025-04-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

##Introduction

In this vignette, we will use lubridate package from the Tidyverse to parse and manipulate data that uses date and time. The data we will use is flights dataset in nycflights13 package. We will calculate departure delays and visualize the results. 

###Load Data and Perform Analysis

First, lets install and load the necessary libraries. 

```{r}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("nycflights13")

library(tidyverse)
library(lubridate)
library(nycflights13)
```

Let's load the flight dataset and get a glimpse of the dataset. 

```{r}
flights <- nycflights13::flights
glimpse(flights)
```

After the glimpse of the dataset, we will use functions in the lubridate package to convert date and time columns into proper datetime objects.

```{r}

flights <- flights %>%
  mutate(
    flight_date = make_date(year, month, day),
    sched_dep_time = str_pad(sched_dep_time, 4, pad = "0"), # fixed string width of 4 otherwise creates problems with sched_departure_time
    dep_time = str_pad(dep_time, 4, pad = "0"),
    sched_departure_time = parse_time(as.character(sched_dep_time), format = "%H%M"), 
    actual_departure_time = parse_time(as.character(dep_time), format = "%H%M")
  )

```

With this, we have 3 new columns in the dataset. The date is combined into one column and the times are combined into one column for easier readability and analysis. 

Next, we will convert the departure delay values into a numeric format for easier analysis. Then we will summarize the average departure delay for each day of the week and create a visualization to present the findings. 

We will use wday function from the lubridate package to figure out what day of the week the date of departure fell on. 

```{r}
#Calculate delay in minutes
flights <- flights %>%
  mutate(departure_delay_minutes = as.numeric(dep_delay))

#Add a day of the week column
flights <- flights %>%
  mutate(day_of_week = wday(flight_date, label = TRUE))

#Summarize average delay by day of the week
average_delay <- flights %>%
  group_by(day_of_week) %>%
  summarize(average_delay = mean(departure_delay_minutes, na.rm = TRUE))

#Plot the results
ggplot(average_delay, aes(x = day_of_week, y = average_delay)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Departure Delay by Day of the Week",
    x = "Day of the Week",
    y = "Average Departure Delay (minutes)"
  )

```

##Conclusion

In this vignette, we utilized the lubridate package to parse and manipulate date and time data in one of the nycflights13 dataset. We calculated departure delays and visualized the average delays by each day of the week. Using lubridate package made it easier to perfrom these analyses. 
