---
title: "Student_depression_vignettes"
author: "W.Durosier"
date: "2025-04-09"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Get Started

## Introduction

For this Vignette, we will be exploring mental health trends among students using th Student Depresion Dataset by April Shamim(Kaggle). The goal is to show how Dplyr and GGplot2-two core components of the TidyVerese- can be utilized to analyze the risk factors accompanied with depression.
Futhermore, will inform and persuade, hilighting ative insight into student well-being and display how TidyVerse tools support transparent, reproducible data anakysis.

## Load Packages

```{r} 
library(tidyverse)
library(dplyr)
library(ggplot2)


```


## Load Data

```{r}

student_data <- c("C:/Users/wdque/Downloads/student_depression_dataset.csv")

student_data <- read.csv("C:/Users/wdque/Downloads/student_depression_dataset.csv")
```

## Exploring data

```{r}
glimpse(student_data)

head(student_data)
```
## CleaN and Transform dataset
 Here the data 'Depression column will be inspected while targeting variables , and prepare categorical variables for analysis.
 
 

```{r}
str(student_data)
```

## Preparing Data
```{r}
# Convert Depression to a factor if it's not already
student_data <- student_data %>%
  mutate(
    Depression = case_when(
      Depression == 1 ~ "Yes",
      Depression == 0 ~ "No",
      TRUE ~ NA_character_
    ),
    Depression = factor(Depression, levels = c("No", "Yes"))
  )

# Check the transformation
table(student_data$Depression, useNA = "ifany")

```
##Analysis Visualization using ggplot2

```{r}
student_data %>%
  ggplot(aes(x = Gender)) +
  geom_bar(fill = "magenta") +
  labs(title = "Gender Distribution", x = "Gender", y = "Count") +
  theme_minimal()


```
## explore Depression by Gender

```{r}
student_data %>%
  filter(!is.na(Depression)) %>%
  ggplot(aes(x = Gender, fill = Depression)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Depression by Gender",
       y = "Proportion", x = "Gender") +
  scale_fill_manual(values = c("No" = "lightgreen", "Yes" = "tomato")) +
  theme_minimal()

```

# Visual Academic Pressure and Depression

```{r}
student_data %>%
  filter(!is.na(Depression)) %>%
  ggplot(aes(x = Depression, y = Academic.Pressure, fill = Depression)) +
  geom_boxplot() +
  labs(title = "Academic Pressure and Depression",
       x = "Depression Status", y = "Academic Pressure") +
  theme_minimal()

```

## Conclusion
This vignette demonstrated how the Tidyverse makes it possible to analyze mental health data in a neat, sophisticated, and repeatable manner. We noticed:

a glaring disparity in the prevalence of depression by gender.

There is typically a correlation between reported depression and increased academic pressure.

These results can assist educators and policymakers in identifying populations that are at risk and in promoting the well-being of students.

***






